---
title: "Creation of arrays"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creation of arrays}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r lib}
library(marray)
```

The basic function to create an array is `marray()` with corresponding dispatcher methods (S3 methods) for different data types like matrix, data.frame and list.

## 1-D array

```{r}
a <- marray(1:11)
a
```

## 2-D array with row-major ordering (C-style)

```{r}
a <- marray(1:12, dim = c(4, 3), order = "C") # order = "C" is default and doesn't need to pass explicitly
a
```

## 2-D array with column-major ordering (F-style)

```{r}
a <- marray(1:12, dim = c(4, 3), order = "F")
a
```

## 3-D array

The reading direction for an n-dimensional array with n \> 2 is generally from right to left. The last dimension, axis is another word for dimension, represents a number of pages, sheets (both terms from [MatLab](https://www.mathworks.com/products/matlab.html){target="_blank"}) or bunches, on each of which there are a number of further bunches. This number is taken from the penultimate dimension. On these bunches, in turn, there are a number of bunches corresponding to the penultimate dimension, and so on. The first two dimensions can then be combined into a matrix.

The following example creates a 4x3x2 array arranged in columns. There are 2 bunches (last dimension), on each of which there is a 4x3 matrix.

```{r}
a <- marray(1:24, dim = c(4, 3, 2), order = "F")
a
```

This example creates a 4x3x2x2 array arranged in rows.

```{r}
a <- marray(seq_len(4*3*2*2), dim = c(4, 3, 2, 2))
a
```

## Explanation of C- and F-ordering style

For better comprehension, the line-by-line arrangement shall be briefly sketched. The C-ordering style ordering increases the last
dimension first, then the second last etc. In opposite, F-ordering style increases the first dimension first, then the second etc.

### C-style

|           |             |             |             |             |
|-----------|-------------|-------------|-------------|-------------|
| **Value** | **1. axis** | **2. axis** | **3. axis** | **4. axis** |
| 1         | 1           | 1           | 1           | 1           |
| 2         | 1           | 1           | 1           | 2           |
| 3         | 1           | 1           | 2           | 1           |
| 4         | 1           | 1           | 2           | 2           |
| 5         | 1           | 2           | 1           | 1           |
| 6         | 1           | 2           | 1           | 2           |
| 7         | 1           | 2           | 2           | 1           |
| 8         | 1           | 2           | 2           | 2           |
| 9         | 1           | 3           | 1           | 1           |
| 10        | 1           | 3           | 1           | 2           |

### F-style

|           |             |             |             |             |
|-----------|-------------|-------------|-------------|-------------|
| **Value** | **1. axis** | **2. axis** | **3. axis** | **4. axis** |
| 1         | 1           | 1           | 1           | 1           |
| 2         | 2           | 1           | 1           | 1           |
| 3         | 3           | 1           | 1           | 1           |
| 4         | 4           | 1           | 1           | 1           |
| 5         | 1           | 2           | 1           | 1           |
| 6         | 2           | 2           | 1           | 1           |
| 7         | 3           | 2           | 1           | 1           |
| 8         | 4           | 2           | 1           | 1           |
| 9         | 1           | 3           | 1           | 1           |
| 10        | 2           | 3           | 1           | 1           |

### Excursus: algorithm for determining the corresponding position within an array

```{r}
shape <- c(4, 3, 2)
axis_index <- rep(1, length(shape))
len_data <- 24
for (d in (seq_len(len_data))) {
  print(axis_index)
  for (i in rev(seq_len(length(shape)))) { # drop rev() for column-major ordering
    axis_index[i] <- axis_index[i] + 1
    if (axis_index[i] <= shape[i])
      break
    else
      axis_index[i] <- 1
  }
}
```

In Python, the algorithm looks like:

```{python}
# shape = (4, 3, 2)
# N = len(shape)
# axis_index = [1, ] * N
# len_data = 24
# for d in range(len_data):
#     print(axis_index)
#     # Get next dimension index
#     # This is how row-major ordering works; without reversed: column-major ordering
#     for i in reversed(range(N)):
#         axis_index[i] += 1
#         if axis_index[i] <= shape[i]:
#             break
#         else:
#             axis_index[i] = 1
```

